Measuring meritocratic beliefs with the ISSP social inequality survey: Possibilities, limitations and alternatives

### 2023-09-07 paper-blinded-RR-IJS.Rmd
- new file paper-blinded-RR-IJS for resubmission to IJS

---
title: "Perceptions and Preferences for Meritocracy in Survey Research"
#css: "custom.css"
linestretch: '1.5'
link-citations: yes
# author: "Juan Carlos Castillo, Julio Iturra, Luis Maldonado, Francisco Meneses & Jorge Atria <br><br>"
abstract: |
  Economic and social inequalities are causing growing concerns and crises in contemporary societies. One factor suggested to explain the persistence of inequality is the belief in meritocracy, which justifies economic disparities based on differences in effort and talent. However, empirical research on meritocracy is relatively new and characterized by diverse conceptualizations and measures, leading to inconsistent findings and interpretations. To address this issue, the present paper proposes a conceptual and measurement framework for perceptions and preferences regarding meritocracy and non-meritocracy, focusing on comparative survey research methodologies. The proposal is first tested through confirmatory factor analysis (CFA) using data from the International Social Survey Programme (ISSP) inequality module, commonly used in comparative research on meritocracy. Based on these findings,  a second study focuses on developing a novel scale that better captures the distinction between perceptions and preferences related to meritocracy. This scale is tested using a sample of 2,141 individuals from Chile. The results of this study contribute to refining the measurement of meritocracy and non-meritocracy and shed light on the importance of considering different dimensions when studying meritocracy in survey research, both conceptually and empirically.
  \newline
  \newline

>>>>

---
title: "Measuring meritocratic beliefs with the ISSP social inequality survey: Possibilities, limitations and alternatives"
#css: "custom.css"
linestretch: '1.5'
link-citations: yes
# author: "Juan Carlos Castillo, Julio Iturra, Luis Maldonado, Francisco Meneses & Jorge Atria <br><br>"
abstract: |
  Economic and social inequalities are causing growing concerns and crises in contemporary societies. One factor suggested to explain the persistence of inequality is the belief in meritocracy, which justifies economic disparities based on differences in effort and talent. A great part of the international research on meritocratic beliefs has used the social inequality module of the ISSP (International Social Survey Programme), which for the first time has offered the possibility of comparing different societies in this regard. However, this research line is still a novel enterprise characterized by diverse conceptualizations and measures, often leading to inconsistent findings and interpretations. Based on a series of studies using ISSP data, the present paper proposes a multidimensional measurement framework for meritocratic beliefs that allows us to classify previous research contributions as well as to identify some vacuums and limitations. Firstly, we test the multidimensional framework for meritocratic beliefs using the ISSP 2009 inequality module through confirmatory factor analysis (CFA) techniques. After identifying the advantages and some limitations of ISSP for a multidiemsional operationalization of meritocratic belifs, in a second study we propose an expanded set of items that better taps the different dimensions of meritocratic beliefs. This measurement instrument was applied to a Chilean sample (N=2,141), offering new evidence regarding the multidimensional character of meritocratic beliefs.
  \newline
  \newline
